Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-05-04 paper.tex
Yan's updates

\subsection{Backbone phylogeny}
For validation purposes, we compare the genetic
relationships in the sc2ts-inferred ARGs with phylogenetic relationships
inferred by the NextStrain project \citep{Hadfield2018-ef}. The NextStrain
trees consist of subsamples of SARS-CoV-2 genomes, which over the time period
of our ARGs are almost entirely restricted to non-recombinant lineages. To
enable this comparison, the ARGs were ``simplified'' \citep{Kelleher2018-xc} to
retain only those samples of our genealogy that match those present in a
specific NextStrain tree. These simplified ``backbone'' ARGs exhibit very
little recombination (6 and 5 trees across the genome for the wide and long
ARGs, respectively).

@@

@@

\includegraphics[width=\textwidth]{figures/Pango_XA_XAG_XD_nxcld_tight_graph.pdf}
% }
\end{tabularx} \caption{\label{fig:pango-simple-origin-graph} Simple origin
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
imputed Pango status. (B) Equivalent plot for the 17 XAG samples in the long ARG
(C) Equivalent plot for both origination events for the four XD samples in the long ARG.
} \end{figure}

@@

\subsubsection{Complex origins} 
\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/Pango_XB_nxcld_tight_graph.pdf}
\caption{\label{fig:complex_origins_graph}  A Pango lineage subgraph showing
complex origins of the recombinant lineage XB. All XB samples in the long ARG
trace to a single recombination node (top centre), but three further recombinations
also descend from this node, with many samples descending from these nested
recombinations not being identified as XB, but instead assigned by Nextclade to
various pre-Alpha lineages.}
\end{figure}

As well as cases where single recombination events explain an X lineage,
such as the examples in Figure~\ref{fig:pango-simple-origin-graph}, many of the
Pango X lineages involve more complex origins,
in which additional recombination events are nested inside the earliest shared
recombination. Frequently, the descendants of the original recombination
also include samples which have not been previously identified as recombinants.
An example is XB in the long ARG,
plotted in Figure~\ref{fig:complex_origins_graph}. Here, a recombination
between a B.1 sample and B.1.627 sample leads both to the XB samples and to a
``hairball'' of further recombination nodes whose descendants have not been
identified as recombinants by Nextclade (blue nodes, ``pre-Alpha'' VoC). However,
in this subgraph, there is only a single sample, identified as XB by Nextclade,
which is descended from these nested recombination events; moreover, this
sample is not identified as XB by GISAID (Supplementary
Figure~\ref{fig:pango_XB_gisaid_graph}). We suggest that investigating cases of
complex origins, and identifying which (if any) of the nested, further
recombination events may be artifactual, is an important area of future
research.

>>>>

\subsection{Backbone phylogeny}

@@

For validation purposes, we compare the genetic
relationships in the sc2ts-inferred ARGs with phylogenetic relationships
inferred by the NextStrain project \citep{Hadfield2018-ef}. The NextStrain
trees consist of subsamples of SARS-CoV-2 genomes, which over the time period
of our ARGs are almost entirely restricted to non-recombinant lineages. To
enable this comparison, the ARGs were ``simplified'' \citep{Kelleher2018-xc} to
retain only those samples of our genealogy that match those present in a
specific NextStrain tree. These simplified ``backbone'' ARGs exhibit very
little recombination (6 and 5 trees across the genome for the wide and long
ARGs, respectively).

@@

\includegraphics[width=\textwidth]{figures/Pango_XA_XAG_XD_nxcld_tight_graph.pdf}
% }
\end{tabularx} \caption{\label{fig:pango-simple-origin-graph} Simple origin
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
imputed Pango status. (B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
(C) Equivalent plot for both origination events involving the four XD samples in the long ARG. Details of the mutations and sample node identities for all three plots are provided in supplementary figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
} \end{figure}

@@

\subsubsection{Complex origins} 
\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/Pango_XB_nxcld_tight_graph.pdf}
\caption{\label{fig:complex_origins_graph}  A Pango lineage subgraph showing
complex origins of the recombinant lineage XB. All XB samples in the long ARG
trace to a single recombination node (top centre), but three further recombinations
also descend from this node. The samples descending from these nested
recombinations include 9 that are assigned by Nextclade to
various non-recombinant pre-Alpha lineages (blue).}
\end{figure}

As well as cases where single recombination events explain an X lineage,
such as the examples in Figure~\ref{fig:pango-simple-origin-graph}, many of the
Pango X lineages involve more complex origins,
in which additional recombination events are nested inside the earliest shared
recombination. Frequently, the descendants of the original recombination
also include samples which have not been previously identified as recombinants.
An example is XB in the long ARG,
plotted in Figure~\ref{fig:complex_origins_graph}. Here, a recombination
between a B.1 sample and B.1.627 sample leads both to the XB samples and to a
``hairball'' of further recombination nodes whose descendants are mostly not
identified as recombinants by Nextclade, and plotted in blue. A similar
pattern is seen when examining XB in the wide ARG (discussion above, plot not shown).
Note that the descendants of nested recombination events include only one
XB sample (pink upper left, with 7 mutations above it); moreover, this sample is
not identified as XB by GISAID (Supplementary Figure~\ref{fig:pango_XB_gisaid_graph})
indicating some uncertainty in lineage assignment in this part of the ARG.
Investigating examples of complex recombinant origins, and identifying which
(if any) of the nested recombination events may be artifactual, is an important
area of future research.
